Read Me

### 2023-04-30 code/analysis_code/1_EDA_alt.Rmd
hopefully this works...

```{r}
pos_predictors <- race_sub %>% 
  select(c("Median Household Income",
           "Percent Living in Poverty",
           "Percent Without an Internet subscription",
           "Percent Uninsured Under 65 Years Old",
           "Percent Living with Food Insecurity",
           "Percentage Diagnosed Late")) %>% 
  names()
```

>>>>

```{r}
pos_predictors <- race_sub %>% 
  select(c("State", 
           "Race",
           "Median Household Income",
           "Percent Living in Poverty",
           "Percent Without an Internet subscription",
           "Percent Uninsured Under 65 Years Old",
           "Percent Living with Food Insecurity",
           "Percentage Diagnosed Late"))
```

### 2023-04-30 code/analysis_code/2_pca_alt.Rmd
hopefully this works...

  #setting up correlation matrix 
print(ggcorrplot(corr_matrix))
ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
```

@@

#Scree plot
print(fviz_eig(pca, addlabels = TRUE))
ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
```
Since we originally had 5 variables, we should only look at components that explain more than 20% of variation. This is includes components 1 through 3. All other principal components explains less that 8.9% of variation, so they will excluded.

```{r}
#Contribution of each variable 
print(fviz_cos2(pca, choice = "var", axes = 1:2))
ggsave(here("results/contribution_plot.png"), units="in", width=10, height=7, dpi=300)

# Biplot of the attributes
print(fviz_pca_var(pca, col.var = "cos2",
                   gradient.cols = c("darkblue", "orange", "green"),
                   repel = TRUE, title = "PCA"
                   ))
ggsave(here("results/biplot.png"), units="in", width=10, height=7, dpi=300)
# Variables in green are a good representation of the variable on that component.

@@

>>>>

  #setting up correlation matrix 
pca_1 <-print(ggcorrplot(corr_matrix))
#ggsave(here("results/corr_mat.png"), units="in", width=10, height=7, dpi=300)
```

@@

#Scree plot
pca_2 <-print(fviz_eig(pca, addlabels = TRUE))
#ggsave(here("results/scree_plot.png"), units="in", width=10, height=7, dpi=300)
```
Since we originally had 5 variables, we should only look at components that explain more than 20% of variation. This is includes components 1 through 3. All other principal components explains less that 8.9% of variation, so they will excluded.

```{r}
#Contribution of each variable 
pca_3 <-print(fviz_cos2(pca, choice = "var", axes = 1:2))
#ggsave(here("results/contribution_plot.png"), units="in", width=10, height=7, dpi=300)

# Biplot of the attributes
pca_4 <- print(fviz_pca_var(pca, col.var = "cos2",
                   gradient.cols = c("darkblue", "orange", "green"),
                   repel = TRUE, title = "PCA"
                   ))
#ggsave(here("results/biplot.png"), units="in", width=10, height=7, dpi=300)
# Variables in green are a good representation of the variable on that component.

@@

```{r}
library(gridExtra)
grid_pca <- grid.arrange(pca_1, pca_2, pca_3, pca_4, ncol = 2, nrow = 2)

ggsave(here("results/grid_pca.png"), grid_pca, units="in", width=10, height=10, dpi=300)
```

### 2023-04-30 code/analysis_code/3_model_alt.Rmd
hopefully this works...

Setting up functions for models
```{r}
 #multiple linear regression
mult_function <- function(df) {
  mult_race <- glm(MRR ~ 
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
  print(summary(mult_race))
    ##pull fitted values
df$mult_fit <- mult_race$fitted.values
return(df)
}

## function to fit GLM and pull AIC
mult_aic <- function(df) {
  mult_race <- glm(MRR ~ 
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
aic <- AIC(mult_race)

@@

pois_race <- glm(`Death Cases` ~ Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
  print(summary(pois_race))

@@

## function to fit GLM and pull AIC
pois_aic <- function(df) {
  pois_race <- glm(`Death Cases` ~  Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
aic <- AIC(pois_race)
return(aic)
}
```

@@

```{r}
  #Negative binomial
nb_function  <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                            data = df)
  print(summary(neg_bin_race))

@@

## function to fit GLM and pull AIC
nb_aic <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old` +
                      #  `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                            data = df)
aic <- AIC(neg_bin_race)

@@

  car_race <- spautolm(MRR ~                      
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))

@@

  car_race <- spautolm(MRR ~                      
                       # `Percent Living in Poverty` +
                        `Gini Coefficient` + 
                        `Percent Without an Internet subscription` +
                        `Percent Uninsured Under 65 Years Old`,
                      #  `Percent Living with Food Insecurity`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
aic <- AIC(car_race)
return(aic)

@@

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$MRR, nablist_race_df)
  summary(lisa_race[,1])
}
```

>>>>

Setting up functions for models
```{r}
 #multiple linear regression
mult_function <- function(df) {
  mult_race <- glm(MRR ~ 
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
  print(summary(mult_race))
    ##pull fitted values
df$mult_fit <- mult_race$fitted.values
return(df)
}

## function to fit GLM and pull AIC
mult_aic <- function(df) {
  mult_race <- glm(MRR ~ 
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity`,
                     data = df,
                     family = "gaussian")
aic <- AIC(mult_race)

@@

pois_race <- glm(`Death Cases` ~ Race + region +
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
  print(summary(pois_race))

@@

## function to fit GLM and pull AIC
pois_aic <- function(df) {
  pois_race <- glm(`Death Cases` ~  Race + region +
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity`  +
                     offset(log(`Population`)),
                     data = df,
                     family = poisson(link = "log"))
aic <- AIC(pois_race)
return(aic)
}
```

@@

```{r}
  #Negative binomial
nb_function  <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity` +
                     offset(log(`Population`)),
                            data = df)
  print(summary(neg_bin_race))

@@

## function to fit GLM and pull AIC
nb_aic <- function(df) {
  neg_bin_race <- glm.nb(`Death Cases` ~ Race + region +
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity` +
                     offset(log(`Population`)),
                            data = df)
aic <- AIC(neg_bin_race)

@@

  car_race <- spautolm(MRR ~                      
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
  print(summary(car_race))

@@

  car_race <- spautolm(MRR ~                      
                        `Median Household Income` +
                        `Percent Living in Poverty` +
                        `Percent Without an Internet subscription` +
                        `Percent Living with Food Insecurity`,
                           data = df,
                           listw = nablist_race_df,
                           family = "CAR")
aic <- AIC(car_race)
return(aic)

@@

  # calculate and print LISA statistics
  lisa_race <- localmoran(df$MRR, nablist_race_df, zero.policy=TRUE)
  summary(lisa_race[,1])
}
```

### 2023-04-30 code/analysis_code/p4_mapping.Rmd
hopefully this works...

```{r}
ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Gini Coefficient`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Gini Coefficient, 2020", x = "", y = "", fill = "Gini coefficient (higher values represent greater income inequality)")  +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/gini_map.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Uninsured Under 65 Years Old, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/uninsured_map.png"), units="in", width=10, height=7, dpi=300)
```

```{r}
ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Without an Internet Subscription, 2020", x = "", y = "", fill = "%") +
  theme(plot.title= element_text(hjust = 0.5)) 
ggsave(here("results/no_internet_map.png"), units="in", width=10, height=7, dpi=300)
```

@@

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_mult)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
United States, 2020 (GLM)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/mult_pred_map.png"), units="in", width=7, height=4, dpi=300)

ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_car)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Map of Predicted Over Observed Black-to-White HIV-Related Mortality Rate Ratio,
United States, 2020 (CAR)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/car_pred_map.png"), units="in", width=7, height=4, dpi=300)
``` 
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values. We can see that the model underestimated rates primarily in the non-coast north-western states and north central states for persons of Hispanic, non-Hispanic White, and non-Hispanic multiple racial/ ethnic backgrounds.

>>>>

```{r}
map_1 <- ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living in Poverty, 2020", x = "", y = "", fill = "Percent")  +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
map_2 <-

  ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Median Household Income`/1e4)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Median Household Income, 2020", x = "", y = "", fill = "Per $10,000")  +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
map_3 <- ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Uninsured Under 65 Years Old, 2020", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
map_4 <- ggplot(df_sf_raw, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Without an Internet Subscription, 2020", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5))

```

```{r}
library(gridExtra)
grid_map<- grid.arrange(map_2, map_1, map_3, map_4, ncol = 2, nrow = 2)

ggsave(here("results/grid_map.png"), grid_map, units="in", width=11, height=10, dpi=300)
```

@@

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
pred_mult <- ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_mult)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Predicted Over Observed
Black-to-White HIV-Related MRR,
United States, 2020 (GLM)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))

pred_car <-ggplot(df_sf_scale, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_car)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Predicted Over Observed
Black-to-White HIV-Related MRR,
United States, 2020 (CAR)", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
```

```{r}
grid_pred <- grid.arrange(pred_car, pred_mult, ncol = 2, nrow = 1)

ggsave(here("results/grid_pred.png"), grid_pred, units="in", width=10, height=5, dpi=300)
```
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values. We can see that the model underestimated rates primarily in the non-coast north-western states and north central states for persons of Hispanic, non-Hispanic White, and non-Hispanic multiple racial/ ethnic backgrounds.
